Demystifying the Computer Science \\ PhD Admission in the US\\{\large A Handbook for International \emph{and Domestic} Students

### 2024-10-28 demystify.tex
updates

\section{Do adcom members read all applications?}\label{sec:read-all}
\sectioninfo{Adcom members only review applications assigned to them (typically matching their expertise) and rarely get involved in other applications}

@@

Some universities do waive, e.g., Rice does for all PhD applicants and many universities for domestic students (\autoref{sec:domestic-students}), however most do not for international students.  In any case, the application fee is typically a requirement of the university. Individual departments and programs do not have the flexibility to waive the application fee, even if they want to.

\begin{commentbox}{Vu:}
In my opinion, requiring applicants to pay the fee helps ensure their seriousness, as it filters out non-serious candidates. Also, most CS programs already receive many applications and would be overwhelmed if the application process were free.  Note that if you have financial difficulties, you can ask the department for a waiver, but this is typically only granted in exceptional cases.
\end{commentbox}

@@

\paragraph{LoR from Famous Ppl?} Having a strong letter from an internationally recognized researcher will \emph{greatly strengthen} your application. A strong LOR from such a researcher can outweigh the lack of publications (or low GPA or unknown school).

@@

\begin{commentbox}[Vu:]
For Vietnamese students, it's worth mentioning the \href{https://vef2.org/}{VEF2.0 program}, which has helped many students gain admission to top CS PhD programs in the US. VEF2.0 invites US faculty members from leading institutions to conduct rigorous interviews with VEF students and subsequently provide reference letters on their behalf. Despite the limited interaction between the interviewer and the interviewee (primarily just the interview), these reference letters are generally effective and have helped many students get into good PhD programs.
\end{commentbox}

\section{Statement of Purpose (SOP)}\label{sec:sop}
\sectioninfo{\emph{SOP is very important}. Write it in such a way that makes you \emph{stand out}.}

@@

While you might not have control over LORs (\autoref{sec:lor}) or where your go to school (\autoref{sec:your-school}), you do over your
statement of purpose (SOP) or personal statement\footnote{Few schools separate these documents and ask you to write both: SOP, which focuses on research experiences, and personal statement, which is everything more personal, e.g., why PhD, challenges, diversity, etc}! So write it well because it could make a big difference. A well-written SOP also shows that you can communicate, which is very important in research, and that you can effectively teach and communicate with students, which is important for TA funding (see \autoref{sec:funding}).

@@

In your SOP, focus on research potential and convince us through your experience, e.g., published papers. You should also back up your claims with concrete evidence. For example, if you say you have experience with teaching, then show what you did, e.g., undergrad TA or mentoring someone.  If you say you work on a project, then show some results, e.g., paper submitted (or even rejected), achieved certain performance improvement over the state of the art. Also, do not use more than two pages (references can be on the third page). Many SOP samples are \href{https://cs-sop.org/}{available here}.

@@

Finally, you should get your SOP reviewed by your LoR writers and professors, especially those who have served in adcom, or even postdocs or PhD students as they have been through this process.

@@

As discussed in \autoref{sec:research-experience}, having a successful research experience can greatly strengthen a PhD application. Research experience gives you opportunities to try out research, determine what research area you're interested in, publish papers (\autoref{sec:research-experience}), connect with researchers, and get strong LoRs (\autoref{sec:lor}).

 This section provides some guidance on how to gain research experience, e.g., as an undergraduate student.  It also advises students from small colleges where research opportunities are limited.

>>>>

\section{How applications are assigned to adcom members?}\label{sec:applications-assigned}
\sectioninfo{Adcom members only review applications assigned to them (typically matching their expertise) and rarely get involved in other applications}

@@

Some universities do waive, e.g., Rice does for all PhD applicants and many universities for domestic students (\autoref{sec:domestic-students}), however most do not for international students.
Some require proof of financial difficulties, e.g., a statement from a financial adviser or a bank statement.

In gneral, application fee is typically a requirement of the university. Individual departments and programs do not have the flexibility to waive the application fee, even if they want to.

\begin{commentbox}{Vu:}
In my opinion, requiring applicants to pay the fee helps ensure their seriousness, as it filters out non-serious candidates. Most CS programs already receive many applications and would be overwhelmed if the application process were free.  Note that if you have financial difficulties, you can ask the department for a waiver, but this is typically only granted in exceptional cases.
\end{commentbox}

@@

\paragraph{LoR from Famous People} Having a strong letter from an internationally recognized researcher will \emph{greatly strengthen} your application. A strong LOR from such a researcher can outweigh the lack of publications (or low GPA or unknown school).

@@

% \begin{commentbox}[Vu:]
% For Vietnamese students, it's worth mentioning the \href{https://vef2.org/}{VEF2.0 program}, which has helped many students gain admission to top CS PhD programs in the US. VEF2.0 invites US faculty members from leading institutions to conduct rigorous interviews with VEF students and subsequently provide reference letters on their behalf. Despite the limited interaction between the interviewer and the interviewee (primarily just the interview), these reference letters are generally effective and have helped many students get into good PhD programs.
% \end{commentbox}

\section{Statement of Purpose (SOP)}\label{sec:sop}
\sectioninfo{\emph{SOP is important}. Write it in such a way that makes you \emph{stand out}.}

@@

While you might not have control over LORs (\autoref{sec:lor}) or where your go to school (\autoref{sec:your-school}), you do over your
statement of purpose (SOP) or personal statement\footnote{Few schools separate these documents and ask you to write both: SOP, which focuses on research experiences, and personal statement, which is everything more personal, e.g., why PhD, challenges, diversity, etc}! A well-written SOP also shows that you can communicate, which is very important in research, and that you can effectively teach and communicate with students, which is important for TA funding (see \autoref{sec:funding}).  Many SOP samples for CS are \href{https://cs-sop.org/}{available here}.

@@

In your SOP, focus on research potential and convince us through your experience, e.g., published papers. Back up your claims with \textbf{concrete evidence}. For example, if you say you have experience with teaching, then show what you did, e.g., undergrad TA or mentoring someone.  If you say you work on a project, then show some results, e.g., paper submitted (or even rejected), achieved certain performance improvement over the state of the art.

@@

Finally, your SOP reviewed by your LoR writers and professors, especially those who have served in adcom, or even postdocs or PhD students as they have been through this process.

@@

As discussed in \autoref{sec:research-experience}, having a successful research experience can greatly strengthen a PhD application. Research experience gives you opportunities to try out research, determine what research area you're interested in, publish papers (\autoref{sec:research-experience}), connect with researchers, and get strong LoRs (\autoref{sec:lor}). This section provides some guidance on how to gain research experience as an undergraduate student or student at a smaller college where research opportunities might be limited.
